Anomaly detection with kernel density estimation on manifolds

### 2022-06-07 paper/paper.tex
Change notation for \hat{f}

Consider a compact Riemannian manifold \((M, g)\) of dimension \(d\) without boundary and a probability distribution with density \(f\) on the manifold. Assume \((M, d_g)\) is a complete metric space, where \(d_g\) is the Riemannian distance induced by the Riemannian, \(g\), and the strictly positive injectivity radius \autocite{Chavel2006-mp} of the manifold, \(inj_g(M)\). Denote \(A_i, i = 1, 2, \dots, N\), where \(\pmb{A}_i \in \mathcal{R}^d\), as i.i.d. random objects on \(M\) with density \(f\). For each \(p\in M\), \textcite{Pelletier2005-vu} proposed the kernel density estimator of \(f\) to be
\begin{equation}
\label{eq:denriem}
f_N(p) = \frac{1}{N} \sum_{i=1}^{N} \frac{1}{h^d \theta_{A_i}(p)} K(\frac{d_g(p, A_i)}{h}),
\end{equation}
where \(h\) is the bandwidth satisfying \(h \leq h_0\) for fixed \(h_0\) such that \(0\leq h_0\leq inj_g(M)\), \(K: \mathcal{R}_{+} \rightarrow \mathcal{R}\) is a non-negative map, and \(\theta_{q}(p)\) is the volume density function on the manifold.
\textcite{Pelletier2005-vu} mention that the volume density function \(\theta_q(p)\) defined for \(p\) in a neighborhood of \(q\) on the manifold, with the geodesic normal coordinates at \(q\), is equal to the determinant of the Riemannian metric \(g\) at the logarithm map at q \(\exp_q^{-1}(p)\). The expression in \eqref{eq:denriem} is also proved to be consistent with the kernel density estimators in the Euclidean case as
\begin{equation}
\label{eq:denconsist}
f_N(p) = \frac{1}{N} \sum_{i=1}^{N} \frac{1}{h^d} K(\frac{\|p-A_i\|}{h}),
\end{equation}
and converges at the same rate as the Euclidean kernel density estimator \autocite{Henry2009-ll}.

However, in application, we usually found the observations lie on a \(d\)-dimensional manifold embedded in a \(s\)-dimensional space, where \(d \ll s\). Then manifold learning can be applies to find the \(d\)-dimensional representation of the manifold \(M\). Therefore, based on the estimator in \eqref{eq:denriem}, we propose to a kernel density estimator of f with the manifold embedding \(y_i, i = 1, 2, \dots, N\) to be
\begin{equation}
\label{eq:denestimator}
f_N(p) = \frac{1}{N} \sum_{i=1}^{N} \frac{1}{h_i^d |g(\exp_p^{-1}(y_i))|} K(\frac{d_g(p, y_i)}{h_i}),
\end{equation}
where \(h_i, i = 1, 2, \dots, N\) is the pointwise variable bandwidth matrix and \(h_i\) is positive definite with dimension
\(d\times d\).

@@

\hypertarget{riemannian-metric-as-variable-bandwidth}{%
\subsection{Riemannian metric as variable bandwidth}\label{riemannian-metric-as-variable-bandwidth}}

@@

\hypertarget{volume-density-function}{%
\subsection{Volume density function}\label{volume-density-function}}

>>>>

Consider a compact Riemannian manifold \((M, g)\) of dimension \(d\) without boundary and a probability distribution with density \(f\) on the manifold. Assume \((M, d_g)\) is a complete metric space, where \(d_g\) is the Riemannian distance induced by the Riemannian, \(g\), and the strictly positive injectivity radius \autocite{Chavel2006-mp} of the manifold, \(inj_g(M)\). Denote \(\pmb{X}_i, i = 1, 2, \dots, N\), where \(\pmb{X}_i \in \mathcal{R}^d\), as i.i.d. random objects on \(M\) with density \(f\). For each point \(p\in M\), \textcite{Pelletier2005-vu} proposed a kernel density estimator of \(f\) to be
\begin{equation}
\label{eq:denriem}
\hat{f}(p) = \frac{1}{N} \sum_{i=1}^{N} \frac{1}{h^d \theta_{\pmb{X}_i}(p)} K(\frac{d_g(p, \pmb{X}_i}{h}),
\end{equation}
where \(h\) is the global bandwidth satisfying \(h \leq h_0\) for fixed \(h_0\) such that \(0\leq h_0\leq inj_g(M)\), \(K: \mathcal{R}_{+} \rightarrow \mathcal{R}\) is a non-negative map, and \(\theta_{q}(p)\) is the volume density function on the manifold.
\textcite{Pelletier2005-vu} mention that the volume density function \(\theta_q(p)\) defined for \(p\) in a neighborhood of \(q\) on the manifold, with the geodesic normal coordinates at \(q\), is equal to the determinant of the Riemannian metric \(g\) at the logarithm map at q, \(\exp_q^{-1}(p)\). The expression in \eqref{eq:denriem} is also proved to be consistent with the kernel density estimators in the Euclidean case as
\begin{equation}
\label{eq:denconsist}
\hat{f}(p) = \frac{1}{N} \sum_{i=1}^{N} \frac{1}{h^d} K(\frac{\|p-\pmb{X}_i\|}{h}),
\end{equation}
and converges at the same rate as the Euclidean kernel density estimator \autocite{Henry2009-ll}.

\hypertarget{volume-density-function}{%
\subsubsection{Volume density function}\label{volume-density-function}}

However, in application, we usually found the observations lie on a \(d\)-dimensional manifold embedded in a \(s\)-dimensional space, where \(d \ll s\). Then manifold learning can be applies to find the \(d\)-dimensional representation of the manifold \(M\). Therefore, based on the estimator in \eqref{eq:denriem}, we propose a kernel density estimator of f with the manifold embedding \(y_i, i = 1, 2, \dots, N\) to be
\begin{equation}
\label{eq:denestimator}
\hat{f}(p) = \frac{1}{N} \sum_{i=1}^{N} \frac{1}{h_i^d |g(\exp_p^{-1}(y_i))|} K(\frac{d_g(p, y_i)}{h_i}),
\end{equation}
where \(h_i, i = 1, 2, \dots, N\) is the pointwise variable bandwidth matrix and \(h_i\) is positive definite with dimension \(d\times d\).

@@

\hypertarget{riemannian-metric-as-variable-bandwidth}{%
\subsubsection{Riemannian metric as variable bandwidth}\label{riemannian-metric-as-variable-bandwidth}}

@@

\hypertarget{volume-density-function-1}{%
\subsection{Volume density function}\label{volume-density-function-1}}
